---
title: "Crossover Trial"
author: "Eamonn O'Brien"
date: '`r format(Sys.time(), "%d %B, %Y")`'
output:
  pdf_document:
    toc: true
    toc_depth: 2
  fig_height: 6
  fig_width: 8
header-includes:
- \usepackage{eso-pic,graphicx,transparent}
- \usepackage{graphicx}
- \usepackage{fancyhdr}
- \pagestyle{fancy}
- \setlength\headheight{22pt}
- \fancyfoot[RO]{Estimating Population Size}
- \usepackage{lastpage}
- \cfoot{Page \thepage\ of \pageref{LastPage}}
---

\newpage  
\tableofcontents  
\listoffigures
\listoftables
\newpage

```{r set-options, echo=FALSE, cache=FALSE, warning = FALSE}

 

         set.seed(123)
         startTime<-proc.time()
         library(knitr)
         options(width=60)

         opts_chunk$set(comment = "", warning = FALSE, message = FALSE,
                       echo = TRUE, tidy = TRUE, size="tiny",  cache=FALSE,
                       progress=TRUE, tidy.opts=list(width.cutoff=60),
                         fig.width=7, fig.height=3.5,
                       cache.path = 'program_Cache/',
                       fig.path='figure/')
         
        # opts_chunk$set(tidy.opts=list(width.cutoff=80),tidy=TRUE) 
         
        knitr::knit_hooks$set(inline = function(x) {
          knitr:::format_sci(x, 'md')
        })
         
        
        options(scipen=999)  # remove scientific notation
        

        
        
 
```

```{r ,echo=FALSE}

p2 <- function(x) {formatC(x, format="f", digits=4)}

```

 

## Simulate a tiny crossover trial to show matrices

```{r function}

 
    n <- 3   # number of patients

    sdB= 1   # between person SD
    
    sdW= 1   # within person SD
    
    beta=c(1, 1, 1, 1)  #intercept, main effect, order effect, interaction
    
    
    Patient   <- as.factor(rep(1:(2*n), rep(2, 2*n)))
    
    Treatment <- c(rep(c("Treatment1", "Treatment2"), n),
                   rep(c("Treatment2", "Treatment1"), n))
    
    Order     <- rep(c("First", "Second"), 2*n)
    
    Data      <- data.frame(Patient, Treatment, Order)
    
    FMat      <- model.matrix(~ Treatment * Order, data=Data)   # Fixed effects
    
    RMat      <- model.matrix(~ 0 + Patient, data=Data)         # Random effects
    
    Response  <- FMat %*% beta + RMat %*% rnorm(2*n, 0, sdB) + rnorm(4*n, 0, sdW) # matrix multiplication
    
    Data$Response <- Response

```

\newpage  

## Show model matrix for fixed effects

```{r fixed}

    FMat
     
  
```

\newpage  

## Show model matrix for random effects

```{r random}

    RMat
     
  
```

\newpage 

## Show the simulated data

```{r data}


    Data 
 
```

\newpage  

## Fit Model

```{r Model}

    (Fit <- lme4::lmer(Response ~ (1 | Patient) + Treatment * Order, data=Data))
    
    confint(Fit)
     
  
```

\newpage  

##  Simulate a very large crossover trial and analyse

```{r larger trial }

    n <- 5000   # number of patients
    sdB= 1   # between person SD
    sdW= 1   # within person SD
    beta=c(1, 1, 1, 1)  #intercept, main effect, order effect, interaction
    Patient   <- as.factor(rep(1:(2*n), rep(2, 2*n)))
    Treatment <- c(rep(c("Treatment1", "Treatment2"), n),
                 rep(c("Treatment2", "Treatment1"), n))
    
    Order     <- rep(c("First", "Second"), 2*n)
    Data      <- data.frame(Patient, Treatment, Order)
    FMat      <- model.matrix(~ Treatment * Order, data=Data)   # Fixed effects
    RMat      <- model.matrix(~ 0 + Patient, data=Data)         # Random effects
    Response  <- FMat %*% beta + RMat %*% rnorm(2*n, 0, sdB) + rnorm(4*n, 0, sdW) # matrix multiplication
    

```

\newpage  

## Analyse the large trial

```{r large trial analysis}
 
    summary(Fit <- lme4::lmer(Response ~ (1 | Patient) + Treatment * Order, data=Data))
    
    confint(Fit)
    
 

  

```
\newpage  

## Computing Environment

```{r}

sessionInfo()

```

```{r echo=FALSE}

stopTime<-proc.time()

```

This took `r (stopTime-startTime)[1][[1]]` seconds to execute.

 