---
title: "Crossover Trial Further Simulation and Plotting"
author: "Eamonn O'Brien"
date: '`r format(Sys.time(), "%d %B, %Y")`'
output:
  pdf_document:
    toc: true
    toc_depth: 2
  fig_height: 6
  fig_width: 8
header-includes:
- \usepackage{eso-pic,graphicx,transparent}
- \usepackage{graphicx}
- \usepackage{fancyhdr}
- \pagestyle{fancy}
- \setlength\headheight{22pt}
- \fancyfoot[RO]{Estimating Population Size}
- \usepackage{lastpage}
- \cfoot{Page \thepage\ of \pageref{LastPage}}
---

\newpage  
\tableofcontents  
\listoffigures
\listoftables
\newpage

```{r set-options, echo=FALSE, cache=FALSE, warning = FALSE}

 

         set.seed(123)
         startTime<-proc.time()
         library(knitr)
         options(width=60)

         opts_chunk$set(comment = "", warning = FALSE, message = FALSE,
                       echo = TRUE, tidy = TRUE, size="tiny",  cache=FALSE,
                       progress=TRUE, tidy.opts=list(width.cutoff=60),
                         fig.width=7, fig.height=3.5,
                       cache.path = 'program_Cache/',
                       fig.path='figure/')
         
        # opts_chunk$set(tidy.opts=list(width.cutoff=80),tidy=TRUE) 
         
        knitr::knit_hooks$set(inline = function(x) {
          knitr:::format_sci(x, 'md')
        })
         
        
        options(scipen=999)  # remove scientific notation
        

```



```{r ,echo=FALSE}

p2 <- function(x) {formatC(x, format="f", digits=4)}

```

## Function to simulate crossover trial dataset

```{r function}
 
  ABBA <- function(n=10, sdW=4, sdB=1, beta=c(8, 1, 0, 0), alpha=0.05) {
    
    # generate data
    Patient   <- as.factor(rep(1:(2*n), rep(2, 2*n)))
    Treatment <- c(rep(c("Treatment1", "Treatment2"), n),
                   rep(c("Treatment2", "Treatment1"), n))
    Order     <- rep(c("First", "Second"), 2*n)
    Data      <- data.frame(Patient, Treatment, Order)
    FMat      <- model.matrix(~ Treatment * Order, data=Data)
    RMat      <- model.matrix(~ 0 + Patient, data=Data)
    Response  <- FMat %*% beta + RMat %*% rnorm(2*n, 0, sdB) + rnorm(4*n, 0, sdW)  # beta here
    Data$Response <- Response
    
    return ( data.frame(  Data))
    # analyse
    # Fit  <- lme(Response~ Treatment * Order, random=~1 | Patient, data=Data, na.action="na.omit" )
    # Est <- fixed.effects(Fit)[2]
    # Ste <- sqrt(vcov(Fit)[2,2])
    # prod(Est + c(-1,1) * qnorm(1-alpha/2) * Ste) > 0  
  }

 

```

\newpage

## Generate data and analyse

```{r analyse}

  mydata <- ABBA(n=10, sdW=2, sdB=2, beta=c(8, 1, 0, 0), alpha=0.05 )
  
  require(nlme)
  f <- lme(Response ~ Treatment * Order, random=~1 | Patient, data=mydata, na.action="na.omit" )
  anova(f)
  summary(f)$tTable

```

\newpage

## Simulation function for crossover design, preparing to vary an input parameter

```{r sim2}

    C.power <- function (a,b,c,d,e,f,g, n.sims=sims){
      
      Treatment <-  rep (NA, n.sims)
      Order <-  rep (NA, n.sims)
      Interaction <-  rep (NA, n.sims)
      
       for (s in 1:n.sims){
        
        fake <-ABBA(n=a, sdW=b, sdB=c, beta=c(d, e, f, g), alpha=0.05 )
        
        possibleError <-  
          tryCatch(f1<-lme(Response~  Treatment * Order, random=~1 | Patient, data=fake,
                           na.action="na.omit" ) , 
                   error=function(e) e)
        
        ###http://stackoverflow.com/questions/8093914
        ###/skip-to-next-value-of-loop-upon-error-in-r-trycatch
        
        if(!inherits(possibleError, "error")){
          
          modelint<-possibleError
          
          z<-as.matrix(summary(modelint)$tTable)
          Treatment[s]   <- z[2,5][[1]]
          Order[s]       <- z[3,5][[1]]
          Interaction[s] <- z[4,5][[1]]
        }}
      
      trt      <- mean(Treatment<0.05)   
      ord    <-  mean(Order<0.05)   
      int     <- mean(Interaction<0.05) 
      
      c( trt, ord, int )
    } 
    
    # stand alone, single example that shows power for 
    # main effect, order effect and interaction effect
    C.power(a=20, b=1,c= 1, d=8, e=1, f=0, g=0, n.sims=100)
     

```

\newpage

## Simulate crossover power and vary between person SD

```{r analyse2}

##execute the functions numerous times assess changing between person SD

    sims <- 200
  
    J <- c( 2,2.5,3,4)  # here we vary
    
    k <- length(J)
    
    dnam = list( betweenSD=J, Estimate=c("trt effect power","order power","interaction power" ))
    
    pwpr <- array( NA, dim=sapply( dnam, length ), dimnames=dnam )
    
    
    system.time(for (i in 1:k)   
    {
      pwpr[i,] <- C.power(a=20, b=2, c=J[i], d=8, e=4, f=0, g=0, n.sims=sims)
    })
    
    print(pwpr, digits=4)
  
   
```

\newpage

## Simulate crossover power and vary within person SD

```{r analyse3}

    sims <- 200
    
    J <- c( 2,2.5,3,4)  # here we vary
    
    k <- length(J)
    
    dnam = list( betweenSD=J, Estimate=c("trt effect power","order power","interaction power" ))
    
    pwpr <- array( NA, dim=sapply( dnam, length ), dimnames=dnam )
    
    
    system.time(for (i in 1:k)   
    {
      pwpr[i,] <- C.power(a=20, b=J[i], c=4, d=8, e=4, f=1, g=0, n.sims=sims)
    })
    
    print(pwpr, digits=4)

```

\newpage


```{r analyse4, eval=FALSE}
  
  mydata <- ABBA(n=10, sdW=2, sdB=2, beta=c(8, 1, 0, 0), alpha=0.05 )
  par(mfrow=c(3,1))
    ggpubr::ggpaired(mydata , x="Treatment", y = "Response",
                   color = "Treatment", line.color = "gray", line.size = 0.4,
                   palette = "npg", title="Pooled across period comparison")+
    stat_compare_means(paired = TRUE)



    ggpubr::ggpaired(mydata[mydata$Order=="First",], x="Treatment", y = "Response",
                  color = "Treatment", line.color = "gray", line.size = 0.4,
                  palette = "npg", , title="First period comparison")+
    stat_compare_means(paired = TRUE)  
    
    ggpubr::ggpaired(mydata[mydata$Order=="Second",], x="Treatment", y = "Response",
                   color = "Treatment", line.color = "gray", line.size = 0.4,
                   palette = "npg" , title="Second period comparison")+
    stat_compare_means(paired = TRUE)
    par(mfrow=c(1,1))


#http://www.surefoss.org/visualisation/visualizing-small-scale-paired-data-combining-boxplots-stripcharts-and-confidence-intervals-in-r/

co.plot<-function (mydata){
  
  par(mfrow=c(2,2))

  mydata$temp <- paste(mydata$Treatment, mydata$Order, sep=".")
  
  pre<-as.vector(mydata[grepl("Treatment1.First", mydata$temp),  4 ])
  post<-as.vector(mydata[grepl("Treatment1.Second", mydata$temp),  4 ])
    s<-seq(length(pre))
  par(bty="l")
  
  boxplot(pre,post,main="Treatment1->Treatment2",xlab="Order",ylab="Sig Score",
          names=c("Treatment1","Treatment2"),col=c("lightgreen","lightblue"))
  stripchart(list(pre,post),vertical=T,pch=16,method="jitter",cex=0.5,add=T)
  segments(rep(0.95,length(pre))[s],pre[s],rep(2,length(pre))[s],post[s],col=1,lwd=0.5)
  
  res<-wilcox.test(post,pre,paired=T,conf.int=T)
  stripchart(post-pre,vertical=T,pch=16,method="jitter",main="Difference",
             ylab="Difference:Treatment2-Treatment1",xlab="Median+/-95%CI")
  points(1,res$estimate,col="red",pch=16,cex=2)
  arrows(1,res$conf.int[1],1,res$conf.int[2],col="red",code=3,lwd=3,angle=90)
  abline(h=0,lty=2)#Zero-effectline
  
  pre<-as.vector(mydata[grepl("Treatment2.First", mydata$temp),  4 ])
  post<-as.vector(mydata[grepl("Treatment2.Second", mydata$temp),  4 ])
  s<-seq(length(pre))
  par(bty="l")
  
  boxplot(pre,post,main="Treatment2->Treatment1",xlab="Order",ylab="Sig Score",
          names=c("Treatment2","Treatment1"),col=c("lightblue","lightgreen"))
  stripchart(list(pre,post),vertical=T,pch=16,method="jitter",cex=0.5,add=T)
  segments(rep(0.95,length(pre))[s],pre[s],rep(2,length(pre))[s],post[s],col=1,lwd=0.5)
  
  res<-wilcox.test(post,pre,paired=T,conf.int=T)
  stripchart(post-pre,vertical=T,pch=16,method="jitter",main="Difference",
             ylab="Difference:Treatment1–Treatment2",xlab="Median+/-95%CI")
  points(1,res$estimate,col="red",pch=16,cex=2)
  arrows(1,res$conf.int[1],1,res$conf.int[2],col="red",code=3,lwd=3,angle=90)
  abline(h=0,lty=2)#Zero-effectline
  #detach(mydata)
  par(mfrow=c(1,1))
  
  }

 x<- ABBA(n=50, sdW=2, sdB=2, beta=c(8, 4, 1, 0), alpha=0.05 )
co.plot( x)


#######################################################################
###########PLOT 1######################################################
#######################################################################


mydata<- ABBA(n=50, sdW=2, sdB=2, beta=c(8, 4, 1, 0), alpha=0.05 )

p4 <- function(x) {formatC(x, format="f", digits=4)}

d1<-mydata
require(reshape)
d1$grp<-paste(d1$Treatment, d1$Order, sep=" ")
d1 <- rename(d1, c(Response="count"))
d1 <- rename(d1, c(grp="spray"))
d1 <- d1[order(d1$spray),]

attach(d1)
sprayTypes <- unique(spray)

y<-as.numeric(as.character(d1[,4]))

plot(y, as.factor(d1[,5]) ,  ylim=c(min(y),max(y)) , xlim=c(1, 5) , xaxt="n",
     main=paste("\nTruth: Intercept (TRT2 mean) ",beta[1],", TRT (main) effect ",beta[2],", \nSecond scan (order) effect ",
                beta[3], ", \n Interaction (TRT only second scan effect)",beta[4],
                ", \n Between subjects SD",sb,", within subjects SD", p4(sw) ),
     xlab="", ylab="Outcome measure", frame.plot  =F , col="white")  
##if not white I get a nasty boxplot

axis(1,  at=1:4, labels=F )

text(x=1:length(sprayTypes)*1.05, par("usr")[3]-0.08 , labels = sprayTypes,
     srt = 45, pos = 2, xpd = TRUE,  
     cex.axis=.5 )

##make colours
chk<-as.character(d1$spray)
x<-as.data.frame(table(chk))
freq<-x[,2]
value<-max(dput(1:dim(x)[1]))
IR<-value+1
clr<-rainbow(IR)
clr<-c("blue","red")
wR<-(2*IR-1:IR)/(2*IR)
##colour made
for (i in 1 : length(sprayTypes)){
  y <- count[spray == sprayTypes[i]]
  n <- sum(spray == sprayTypes[i])
  points(jitter(rep(i, n), amount = .1), y, pch = 16, cex = .5,
         
         
         col=ifelse(i<3,clr[1],clr[2])
         
  )
  
  col=ifelse(Treatment=="Treatment1",clr[1],clr[2])
  
  lines(i + c(.12, .28), rep(mean(y), 2), lwd = 1, col="black")
  lines(rep(i + .2, 2),
        mean(y) + c(-1.96, 1.96) * sd(y) / sqrt(n),  lwd = 1, col="black"
        
  )
}

detach(d1)
#######################################################################
#######################################################################
#######################################################################

```

\newpage

## Function to simulate crossover trial using nlme 

```{r nlme function}

 
  

```
\newpage  

## Computing Environment

```{r}

sessionInfo()

```

```{r echo=FALSE}

stopTime<-proc.time()

```

This took `r (stopTime-startTime)[1][[1]]` seconds to execute.

 