---
title: "Crossover Trial Further Simulation and Plotting"
author: "Eamonn O'Brien"
date: '`r format(Sys.time(), "%d %B, %Y")`'
output:
  pdf_document:
    toc: true
    toc_depth: 2
  fig_height: 6
  fig_width: 8
header-includes:
- \usepackage{eso-pic,graphicx,transparent}
- \usepackage{graphicx}
- \usepackage{fancyhdr}
- \pagestyle{fancy}
- \setlength\headheight{22pt}
- \fancyfoot[RO]{Crossover study design}
- \usepackage{lastpage}
- \cfoot{Page \thepage\ of \pageref{LastPage}}
---

\newpage  
\tableofcontents  
\listoffigures
\listoftables
\newpage

```{r set-options, echo=FALSE, cache=FALSE, warning = FALSE}

 

         set.seed(123)
         startTime<-proc.time()
         library(knitr)
         options(width=60)

         knitr::opts_chunk$set(dev = 'pdf') # helps with plots
         
         opts_chunk$set(comment = "", warning = FALSE, message = FALSE,
                       echo = TRUE, tidy = TRUE, size="tiny",  cache=FALSE,
                       progress=TRUE, tidy.opts=list(width.cutoff=60),
                         fig.width=7, fig.height=3.5,
                       cache.path = 'program_Cache/',
                       fig.path='figure/')
         
        # opts_chunk$set(tidy.opts=list(width.cutoff=80),tidy=TRUE) 
         
        knitr::knit_hooks$set(inline = function(x) {
          knitr:::format_sci(x, 'md')
        })
         
        
        options(scipen=999)  # remove scientific notation
        

```



```{r ,echo=FALSE}

p2 <- function(x) {formatC(x, format="f", digits=2)}
p4 <- function(x) {formatC(x, format="f", digits=4)}

```

## Function to simulate crossover trial dataset

```{r function}
 
  ABBA <- function(n=10, sdW=4, sdB=1, beta=c(8, 1, 0, 0), alpha=0.05) {
    
    # generate data
    Patient   <- as.factor(rep(1:(2*n), rep(2, 2*n)))
    Treatment <- c(rep(c("Treatment1", "Treatment2"), n),
                   rep(c("Treatment2", "Treatment1"), n))
    Order     <- rep(c("First", "Second"), 2*n)
    Data      <- data.frame(Patient, Treatment, Order)
    FMat      <- model.matrix(~ Treatment * Order, data=Data)
    RMat      <- model.matrix(~ 0 + Patient, data=Data)
    Response  <- FMat %*% beta + RMat %*% rnorm(2*n, 0, sdB) + rnorm(4*n, 0, sdW)  # beta here
    Data$Response <- Response
    
    return ( data.frame(  Data))
    # analyse
    # Fit  <- lme(Response~ Treatment * Order, random=~1 | Patient, data=Data, na.action="na.omit" )
    # Est <- fixed.effects(Fit)[2]
    # Ste <- sqrt(vcov(Fit)[2,2])
    # prod(Est + c(-1,1) * qnorm(1-alpha/2) * Ste) > 0  
  }

 

```

\newpage

## Generate data and analyse

```{r analyse}

  mydata <- ABBA(n=10, sdW=2, sdB=2, beta=c(8, 1, 0, 0), alpha=0.05 )
  
  require(nlme)
  f <- lme(Response ~ Treatment * Order, random=~1 | Patient, data=mydata, na.action="na.omit" )
  anova(f)
  summary(f)$tTable

```

\newpage

## Simulation function for crossover design, preparing to vary an input parameter

```{r sim2}

    C.power <- function (a,b,c,d,e,f,g, n.sims=sims){
      
      Treatment <-  rep (NA, n.sims)
      Order <-  rep (NA, n.sims)
      Interaction <-  rep (NA, n.sims)
      
       for (s in 1:n.sims){
        
        fake <-ABBA(n=a, sdW=b, sdB=c, beta=c(d, e, f, g), alpha=0.05 )
        
        possibleError <-  
          tryCatch(f1<-lme(Response~  Treatment * Order, random=~1 | Patient, data=fake,
                           na.action="na.omit" ) , 
                   error=function(e) e)
        
        ###http://stackoverflow.com/questions/8093914
        ###/skip-to-next-value-of-loop-upon-error-in-r-trycatch
        
        if(!inherits(possibleError, "error")){
          
          modelint<-possibleError
          
          z<-as.matrix(summary(modelint)$tTable)
          Treatment[s]   <- z[2,5][[1]]
          Order[s]       <- z[3,5][[1]]
          Interaction[s] <- z[4,5][[1]]
        }}
      
      trt      <- mean(Treatment<0.05)   
      ord    <-  mean(Order<0.05)   
      int     <- mean(Interaction<0.05) 
      
      c( trt, ord, int )
    } 
    
    # stand alone, single example that shows power for 
    # main effect, order effect and interaction effect
    C.power(a=20, b=1,c= 1, d=8, e=1, f=0, g=0, n.sims=100)
     

```

\newpage

## Simulate crossover power and vary between person SD

```{r analyse2}

##execute the functions numerous times assess changing between person SD

    sims <- 200
  
    J <- c( 2,2.5,3,4)  # here we vary
    
    k <- length(J)
    
    dnam = list( betweenSD=J, Estimate=c("trt effect power","order power","interaction power" ))
    
    pwpr <- array( NA, dim=sapply( dnam, length ), dimnames=dnam )
    
    
    system.time(for (i in 1:k)   
    {
      pwpr[i,] <- C.power(a=20, b=2, c=J[i], d=8, e=4, f=0, g=0, n.sims=sims)
    })
    
    print(pwpr, digits=4)
  
   
```

\newpage

## Simulate crossover power and vary within person SD

```{r analyse3}

    sims <- 200
    
    J <- c( 2,2.5,3,4)  # here we vary
    
    k <- length(J)
    
    dnam = list( betweenSD=J, Estimate=c("trt effect power","order power","interaction power" ))
    
    pwpr <- array( NA, dim=sapply( dnam, length ), dimnames=dnam )
    
    
    system.time(for (i in 1:k)   
    {
      pwpr[i,] <- C.power(a=20, b=J[i], c=4, d=8, e=4, f=1, g=0, n.sims=sims)
    })
    
    print(pwpr, digits=4)

```

\newpage

## Simulate a ABBA data set with known effects that has good power

```{r sim data, eval=TRUE}

    #require(ggpubr)
    require(ggplot2)

    # state some parameters so we can pick them up later on in plot titles
    sb <- 2
    sw <- 2
    beta=c(8, 2, 1, 0)
    
    #check the power, it is not likely the main effect will be missed, we even have good power for the order effect.
    C.power(a=100, b=sb, c=sw, d=beta[1], e=beta[2], f=beta[3], g=beta[4], n.sims=1000)
     
    # so lets create the data
    mydata <-  ABBA(n=100, sdW=2, sdB=2, beta=c(8, 2, 1, 0), alpha=0.05 )
     
    # analysis
    tryCatch(f1<-lme(Response~  Treatment * Order, random=~1 | Patient, data=mydata,
                           na.action="na.omit" ) , 
                   error=function(e) e)
    
    summary(f1)
    intervals(f1)   
    
    # useful concatenation for plotting
    mydata$temp <- paste(mydata$Treatment, mydata$Order, sep=".")

```

\newpage

## Summary stats

```{r stats, eval=TRUE}    
    
    options(digits = 3)
    library("tidyverse")
    group_by(mydata, temp) %>%
      summarise(
        count = n(),
        mean=mean(Response, na.rm = TRUE),
        SD = sd(Response, na.rm = TRUE),
        median = median(Response, na.rm = TRUE),
        IQR = IQR(Response, na.rm = TRUE)
      )

    #
    #
    # I tried this package, it was only ok
  
    # B <- mydata[mydata$temp %in% "Treatment1.First" | mydata$temp %in% "Treatment2.Second",]
    # 
    # B1  <- ggpubr::ggpaired(B, x="Treatment", y = "Response",
    #               color = "Treatment", line.color = "gray", line.size = 0.4,
    #               palette = "npg", , title="First period comparison") +
    # stat_compare_means(paired = TRUE)  
    # 
    # C <- mydata[mydata$temp %in% "Treatment2.First" | mydata$temp %in% "Treatment1.Second",]
    # 
    # 
    # C1 <- ggpubr::ggpaired(C, x="Treatment", y = "Response",
    #                color = "Treatment", line.color = "gray", line.size = 0.4,
    #                palette = "npg" , title="Second period comparison")+
    # stat_compare_means(paired = TRUE)
   
    # print(B1);print(C1)
     
```

\newpage

## Use a package to plot AB sequence then BA sequence, should be mirror image ideally

```{r plot1, eval=TRUE, results='asis'}    
    
    require("PairedData")
    Treatment1.First <- subset(mydata,  temp == "Treatment1.First", Response, drop = TRUE)
    Treatment2.Second <- subset(mydata,  temp == "Treatment2.Second", Response, drop = TRUE)
     
    pd <- paired(Treatment1.First, Treatment2.Second)
    plot(pd, type = "profile") + theme_bw()
       
    cat("\n\n\\pagebreak\n")
 
    Treatment2.First <- subset(mydata,  temp == "Treatment2.First", Response, drop = TRUE)
    Treatment1.Second <- subset(mydata,  temp == "Treatment1.Second", Response, drop = TRUE)
     
    pd <- paired(Treatment2.First, Treatment1.Second)
    plot(pd, type = "profile") + theme_bw()
           


```

\newpage

## My code to plot treatment comparisons

```{r plot2 code, fig.width=14, fig.height=16,    eval=TRUE} 

#http://www.surefoss.org/visualisation/visualizing-small-scale-paired-data-combining-boxplots-stripcharts-and-confidence-intervals-in-r/

   # main effect 4 and an order effect of 1, no interaction
   # so second period gets a boost of +1

 
  co.plot<-function (mydata){
    
    par(mfrow=c(2,2))
  
    # creat a concatenated variable
    mydata$temp <- paste(mydata$Treatment, mydata$Order, sep=".")
    
    # pull out AB data
    post <- as.vector(mydata[grepl("Treatment1.First",  mydata$temp),  4 ])
    pre <-  as.vector(mydata[grepl("Treatment2.Second", mydata$temp),  4 ])
    s <- seq(length(pre))
    par(bty="l")
    
    # boxplot AB
    boxplot(post,pre,main="Treatment1 -> Treatment2 (AB)",xlab="Order",ylab="Response",
            names=c("Period 1 (TRT1)","Period 2 (TRT2)"),col=c("lightgreen","lightblue"))
    
    # Add raw data with jitter
    stripchart(list(post,pre),vertical=T,pch=16,method="jitter",cex=0.5,add=T)
    # join lines, this would be better if jittered points were joined
    segments(rep(0.95,length(pre))[s],post[s],rep(2,length(pre))[s],pre[s],col=1,lwd=0.5)
    
    # Wilcoxon signed rank test 
    res <- wilcox.test(post,pre,paired=T,conf.int=T)
    stripchart(post-pre,vertical=T,pch=16,method="jitter",main="Median of all differences TRT1 - TRT2",
               ylab="Difference:Treatment1-Treatment2",xlab="Median of all differences +/-95%CI")
    points(1,res$estimate,col="red",pch=16,cex=2)
    arrows(1,res$conf.int[1],1,res$conf.int[2],col="red",code=3,lwd=3,angle=90)
    abline(h=0,lty=2)#Zero-effect line
    
    
    
    
    
    
    
    
    # repeat for BA sequence
    pre<-as.vector(mydata[grepl("Treatment2.First", mydata$temp),  4 ])
    post<-as.vector(mydata[grepl("Treatment1.Second", mydata$temp),  4 ])
    s<-seq(length(pre))
    par(bty="l")
    
    boxplot(pre,post,main="Treatment2 -> Treatment1 (BA)",xlab="Order",ylab="Response",
            names=c("Period 1 (TRT2)","Period 2 (TRT1)"),col=c("lightblue","lightgreen"))
    stripchart(list(pre,post),vertical=T,pch=16,method="jitter",cex=0.5,add=T)
    segments(rep(0.95,length(pre))[s],pre[s],rep(2,length(pre))[s],post[s],col=1,lwd=0.5)
    
    res<-wilcox.test(post,pre,paired=T,conf.int=T)
    stripchart(post-pre,vertical=T,pch=16,method="jitter",main="Median of all differences TRT1 - TRT2",
               ylab="Difference:Treatment1–Treatment2",xlab="Median of all differences +/-95%CI")
    points(1,res$estimate,col="red",pch=16,cex=2)
    arrows(1,res$conf.int[1],1,res$conf.int[2],col="red",code=3,lwd=3,angle=90)
    abline(h=0,lty=2) # Zero-effect line
    par(mfrow=c(1,1))
    
    }
  
```

\newpage

## Plot AB sequence then BA sequence, should be mirror image ideally


```{r plot4 , fig.width=14, fig.height=16,    eval=TRUE} 
  
   co.plot(mydata)

```

\newpage

## My code to plot all data in a useful way

```{r plot5,  fig.width=14, fig.height=10,  eval=TRUE, results='asis'} 


      d1 <- mydata
      require(reshape)
      d1$grp <- paste(d1$Treatment, d1$Order, sep=" ")
      d1 <- rename(d1, c(Response="count"))
      d1 <- rename(d1, c(grp="spray"))
      d1 <- d1[order(d1$spray),]
      
      attach(d1)
      sprayTypes <- unique(spray)
      
      y <- as.numeric(as.character(d1[,4]))
      
      plot(y, as.factor(d1[,5]) ,  ylim=c(min(y),max(y)) , xlim=c(1, 5) , xaxt="n",
           main=paste("\nTruth: Intercept ",beta[1],", Treatment (main) effect ",beta[2],", order effect ", beta[3], ", interaction (TRT only second order effect)",beta[4],
                      ", \n Between subjects SD",p2(sb),", within subjects SD", p2(sw) ),
           xlab="", ylab="Outcome measure", frame.plot  =F , col="white")  
      ##if not white I get a nasty boxplot
      
      axis(1,  at=1:4, labels=F )
      
      text(x=1:length(sprayTypes)*1.05, par("usr")[3]-0.08 , labels = sprayTypes,
           srt = 45, pos = 2, xpd = TRUE,  
           cex.axis=.5 )
      
      ##make colours
      chk <- as.character(d1$spray)
      x <- as.data.frame(table(chk))
      freq <- x[,2]
      value <- max(dput(1:dim(x)[1]))
      IR <- value+1
      clr <- rainbow(IR)
      clr <- c("blue","red")
      wR <- (2*IR-1:IR)/(2*IR)
      ##colour made
      for (i in 1 : length(sprayTypes)){
        y <- count[spray == sprayTypes[i]]
        n <- sum(spray == sprayTypes[i])
        points(jitter(rep(i, n), amount = .1), y, pch = 16, cex = .5,
              col=ifelse(i<3,clr[1],clr[2])
               
        )
        
        col=ifelse(Treatment=="Treatment1",clr[1],clr[2])
        
        lines(i + c(.12, .28), rep(mean(y), 2), lwd = 1, col="black")
        lines(rep(i + .2, 2),
              mean(y) + c(-1.96, 1.96) * sd(y) / sqrt(n),  lwd = 1, col="black"
              
        )
      }

      
 
```


```{r comment,  echo=FALSE, results='asis'}    

    cat("\nThis is a useful plot to look at all the data together. Perhaps lines joining the pairs might be an improvement to show between and within person variation. The crosses are the arithmetic means and 95% CIs. The first left cross is an estimate of the intercept. We can see the treatment effect, the red data is greater than blue. The order effect is apparent too, second period is higher than first. We are expecting this (we know the truth). We did power this study, so there is a high probability the main effect should manifest and to a lesser probability also the order. We know in truth there is no interaction.\n") 
     

```
\newpage  

## Computing Environment

```{r}

sessionInfo()

```

```{r echo=FALSE}

stopTime<-proc.time()

```

This took `r (stopTime-startTime)[1][[1]]` seconds to execute.

 