---
title: "Crossover Trial Power using Simulation"
author: "Eamonn O'Brien"
date: '`r format(Sys.time(), "%d %B, %Y")`'
output:
  pdf_document:
    toc: true
    toc_depth: 2
  fig_height: 6
  fig_width: 8
header-includes:
- \usepackage{eso-pic,graphicx,transparent}
- \usepackage{graphicx}
- \usepackage{fancyhdr}
- \pagestyle{fancy}
- \setlength\headheight{22pt}
- \fancyfoot[RO]{Estimating Population Size}
- \usepackage{lastpage}
- \cfoot{Page \thepage\ of \pageref{LastPage}}
---

\newpage  
\tableofcontents  
\listoffigures
\listoftables
\newpage

```{r set-options, echo=FALSE, cache=FALSE, warning = FALSE}

 

         set.seed(123)
         startTime<-proc.time()
         library(knitr)
         options(width=60)

         opts_chunk$set(comment = "", warning = FALSE, message = FALSE,
                       echo = TRUE, tidy = TRUE, size="tiny",  cache=FALSE,
                       progress=TRUE, tidy.opts=list(width.cutoff=60),
                         fig.width=7, fig.height=3.5,
                       cache.path = 'program_Cache/',
                       fig.path='figure/')
         
        # opts_chunk$set(tidy.opts=list(width.cutoff=80),tidy=TRUE) 
         
        knitr::knit_hooks$set(inline = function(x) {
          knitr:::format_sci(x, 'md')
        })
         
        
        options(scipen=999)  # remove scientific notation
        

```

## 

```{r ,echo=FALSE}

p2 <- function(x) {formatC(x, format="f", digits=4)}

```

## Function to simulate crossover trial using lme4 (warning mnessage: boundary (singular) fit: see ?isSingular)

```{r function}

 
# n - number of patients in each order group
# sdW - within patient standard deviation
# sdB - between patient standard deviation
# beta - coefficient vector c(Intercept, Treatment, Order, Treatment:Order)

    require("lme4")

  simulatec <- function(n=10, sdW=4, sdB=1, beta=c(8, 1, 0, 0), alpha=0.05) {
  
    # generate data
    Patient   <- as.factor(rep(1:(2*n), rep(2, 2*n)))
    Treatment <- c(rep(c("Treatment1", "Treatment2"), n),
                   rep(c("Treatment2", "Treatment1"), n))
    Order     <- rep(c("First", "Second"), 2*n)
    Data      <- data.frame(Patient, Treatment, Order)
    FMat      <- model.matrix(~ Treatment * Order, data=Data)
    RMat      <- model.matrix(~ 0 + Patient, data=Data)
    Response  <- FMat %*% beta + RMat %*% rnorm(2*n, 0, sdB) + rnorm(4*n, 0, sdW)
    Data$Response <- Response
    
    # analyse
    Fit <- lmer(Response ~ (1 | Patient) + Treatment * Order, data=Data)
    Est <- fixef(Fit)[2]
    Ste <- sqrt(vcov(Fit)[2,2])
    prod(Est + c(-1,1) * qnorm(1-alpha/2) * Ste) > 0
 } 

  # power for n=20 and treatment effect 4 
   mean(replicate(1000, simulatec(n=20, beta=c(8, 4, 0, 0))))

```

\newpage

## Function to simulate crossover trial using nlme (no warning messages for equivalent analyses compared to lme4)

```{r nlme function}

# A random effect very near zero seems to be the reason for the warning message using lme4 

 require(nlme)

 simulaten <- function(n=10, sdW=4, sdB=1, beta=c(8, 1, 0, 0), alpha=0.05) {
  
    # generate data
    Patient   <- as.factor(rep(1:(2*n), rep(2, 2*n)))
    Treatment <- c(rep(c("Treatment1", "Treatment2"), n),
                   rep(c("Treatment2", "Treatment1"), n))
    Order     <- rep(c("First", "Second"), 2*n)
    Data      <- data.frame(Patient, Treatment, Order)
    FMat      <- model.matrix(~ Treatment * Order, data=Data)
    RMat      <- model.matrix(~ 0 + Patient, data=Data)
    Response  <- FMat %*% beta + RMat %*% rnorm(2*n, 0, sdB) + rnorm(4*n, 0, sdW)
    Data$Response <- Response
    
    # analyse
    Fit  <- lme(Response~ Treatment * Order, random=~1 | Patient, data=Data, na.action="na.omit" )
    Est <- fixed.effects(Fit)[2]
    Ste <- sqrt(vcov(Fit)[2,2])
    prod(Est + c(-1,1) * qnorm(1-alpha/2) * Ste) > 0 #(logic: neg*neg = pos, so excludes 0)
}
       
 # power for n=20 and treatment effect 4 
   mean(replicate(1000, simulaten(n=20, beta=c(8, 4, 0, 0))))

  
```

\newpage  

## Show model matrix for fixed effects

```{r fixed}
 
  
```

\newpage  

## Show model matrix for random effects

```{r random}
 
     
  
```

\newpage 

## Show the simulated data

```{r data}
 
```

\newpage  

## Fit Model

```{r Model}
 
  
```

\newpage  

##  Simulate a very large crossover trial and analyse

```{r larger trial }
 
    

```

\newpage  

## Analyse the large trial

```{r large trial analysis}
 
     

  

```
\newpage  

## Computing Environment

```{r}

sessionInfo()

```

```{r echo=FALSE}

stopTime<-proc.time()

```

This took `r (stopTime-startTime)[1][[1]]` seconds to execute.

 