---
title: "Crossover Trial Power Simulation en mass and Plotting"
author: "Eamonn O'Brien"
date: '`r format(Sys.time(), "%d %B, %Y")`'
output:
  pdf_document:
  toc: true
toc_depth: 2
fig_height: 6
fig_width: 8
header-includes:
- \usepackage{eso-pic,graphicx,transparent}
- \usepackage{graphicx}
- \usepackage{fancyhdr}
- \pagestyle{fancy}
- \setlength\headheight{22pt}
- \fancyfoot[RO]{Crossover study power}
- \usepackage{lastpage}
- \cfoot{Page \thepage\ of \pageref{LastPage}}
---
  

\tableofcontents  
\listoffigures
\listoftables
\newpage

```{r set-options, echo=FALSE, cache=FALSE, warning = FALSE}

         set.seed(123)
         startTime<-proc.time()
         library(knitr)
         options(width=60)

         knitr::opts_chunk$set(dev = 'pdf') # helps with plots
         
         opts_chunk$set(comment = "", warning = FALSE, message = FALSE,
                       echo = TRUE, tidy = TRUE, size="tiny",  cache=FALSE,
                       progress=TRUE, tidy.opts=list(width.cutoff=60),
                         fig.width=7, fig.height=3.5,
                       cache.path = 'program_Cache/',
                       fig.path='figure/')
         
        # opts_chunk$set(tidy.opts=list(width.cutoff=80),tidy=TRUE) 
         
        knitr::knit_hooks$set(inline = function(x) {
          knitr:::format_sci(x, 'md')
        })
         
        
        options(scipen=999)  # remove scientific notation
        

```



```{r ,echo=FALSE}

p2 <- function(x) {formatC(x, format="f", digits=2)}
p4 <- function(x) {formatC(x, format="f", digits=4)}

```

## Function to simulate crossover trial dataset including simulating missing data

```{r function}

# 'miss' is the percentage of data that is missing
# Entering n=10 for example will mean there are a total of 20 patients, 10 in each treatment sequence
# So n is the number in each treatment sequence

ABBA <- function(n, intercept, main.effect, order.effect, interaction, sdb, sdw, miss){
  
  beta <- c(intercept, main.effect, order.effect, interaction)
  Patient   <- as.factor(rep(1:(2*n), rep(2, 2*n)))
  Treatment <- c(rep(c("Treatment1", "Treatment2"), n),
                 rep(c("Treatment2", "Treatment1"), n))
  Order     <- rep(c("First", "Second"), 2*n)
  Data      <- data.frame(Patient, Treatment, Order)
  FMat      <- model.matrix(~ Treatment * Order, data=Data)
  RMat      <- model.matrix(~ 0 + Patient, data=Data)
  Response  <- FMat %*% beta + RMat %*% rnorm(2*n, 0, sdb) + rnorm(4*n, 0, sdw)  
  Data$Response <- Response
  df <- as.data.frame(Data)
  df$Response[sample(nrow(df), round( nrow(df)*miss ,0) )] <- NA  # may set some data to NA here
  return ( data.frame(df))
  
}


```

\newpage  

## Show a simulated data sets


```{r }

ABBA(n=10, intercept=1, main.effect=0, order.effect=0, interaction=0, sdb=1, sdw=1, miss=.1) # show a simulated data set

```

\newpage  

## Function to use prior function repeatedly

```{r function2}

require(nlme)

IHC.power <- function (a,b,c,d,e,f,g, n.sims=sims, miss) {
  
  treatment <- order <- interaction <- rep(NA, n.sims)  # capture output
  
  for (s in 1:n.sims) {
    
    # create a data set using above function
    fake <- ABBA(n=a, intercept=b, main.effect=c, order.effect=d, interaction=e, sdb=f, sdw=g, miss=miss)
    
    # analyse
    possibleError <-  
      tryCatch(lme(Response ~  Treatment * Order , random =~1 | Patient, data=fake,
                   na.action="na.omit" ) , 
               error=function(e) e)
    
    ###http://stackoverflow.com/questions/8093914
    ###/skip-to-next-value-of-loop-upon-error-in-r-trycatch
    
    if(!inherits(possibleError, "error")){
      
      modelint <- possibleError
      
      z <- as.matrix(summary(modelint)$tTable)
      treatment[s]    <- z[2,5][[1]]
      order[s]        <- z[3,5][[1]]
      interaction[s]  <- z[4,5][[1]]
      
    }}
  
  A     <- mean(treatment   < 0.05)   
  B     <- mean(order       < 0.05)   
  C     <- mean(interaction < 0.05) 
  
  c( A, B, C )
  
} 


```

\newpage  

## Function to use function repeatedly and vary the inputs, print power 

```{r function3}

# execute the functions numerous times varying inputs
sims <- 1000

# these are the parameters that vary 
n <-   c(10, 20, 30, 40 ,50)   # double to find the actual no. of patients
m <-   c(0, 0.10, 0.25)        # missing data %
sdw <- c(1.5, 2.0)             # within person sd


# set up array to capture output
dnam = list(  N=n, Miss.perc= m, SD.within=sdw , Power=c("main.effect","order","interaction" ))
pwpr <- array( NA, dim=sapply( dnam, length ), dimnames=dnam )
str(pwpr)

# run the simulations, set up the truth: n; intercept; ;main.effect; order.effect; interaction; sdb; sdw=; miss;
system.time(
  
  for (i in 1:length(n))
    for (j in 1:length(m))
      for (k in 1:length(sdw)) 
        
        pwpr[i,j,k,] <- IHC.power( a=n[i], b=8, c=.9, d=0, e=0, f=1, g=sdw[k], miss=m[j], n.sims=sims )
  
)

```

\newpage  

## Print power for scenarios

```{r function4}

print(pwpr, digits=4)


```

\newpage  

## Plot power in which proportion of missing data effect on power (main/order/interaction) is explored

```{r function5}


# hold within SD constant at first value cycle through main, order and interaction
lR <- length(m)  # this will vary in each plot
clr <- rainbow( lR ) 
wR <- (2*lR-1:lR)/(2*lR) 


par(mfrow=c(1,3)) 
for( wh in dimnames(pwpr)[[4]] )  # (pwpr[,,"1",wh]),
  matplot( n, (pwpr[,,"1.5",wh]),  # within person SD held constant
           type="l", lty=1, lwd=3, col=clr, 
           ylab=paste0(wh," power"), xlim=c(0,max(n)*1.1), bty="n", las=1 ) 

text( par("usr")[rep(2,lR)], 
      par("usr")[4]*wR + par("usr")[3]*(1-wR), 
      dimnames(pwpr)[[2]], adj=1, col=clr )

```

\newpage  

## Plot power in which within person SD effect on power (main/order/interaction) is explored

```{r function6}


# hold missing constant at first value cycle through main, order and interaction
lR <- length(sdw)  # this will vary in each plot
clr <- rainbow( lR ) 
wR <- (2*lR-1:lR)/(2*lR)


par(mfrow=c(1,3)) 
for( wh in dimnames(pwpr)[[4]] )  # (pwpr[,,"1",wh]),
  matplot( n, (pwpr[,"0",,wh]),   # missing % held constant
           type="l", lty=1, lwd=3, col=clr, 
           ylab=paste0(wh," power"), xlim=c(0,max(n)*1.1), bty="n", las=1 ) 

text( par("usr")[rep(2,lR)], 
      par("usr")[4]*wR + par("usr")[3]*(1-wR), 
      dimnames(pwpr)[[3]], adj=1, col=clr )


```

## References 

 http://www.r-bloggers.com/statistical-aspects-of-two-way-cross-over-studies/  
 http://stackoverflow.com/questions/8289463/extracting-fixed-effects-and-standard-errors-from-several-lme-objects-in-r  
 http://www.r-bloggers.com/simulated-powerprecision-analysis/  
 http://biostatmatt.com/archives/2315  


```{r comment,  echo=FALSE, results='asis'}    

    cat("\n\n") 
     

```

\newpage  

## Computing Environment

```{r}

sessionInfo()

```


```{r echo=FALSE}

stopTime<-proc.time()

```

This took `r (stopTime-startTime)[1][[1]]` seconds to execute.


