---
title: "THE TWO-PERIOD CROSS-OVER CLINICAL TRIAL"
author: "Eamonn O'Brien"
date: '`r format(Sys.time(), "%d %B, %Y")`'
output:
  pdf_document:
    toc: true
    toc_depth: 2
  fig_height: 6
  fig_width: 8
header-includes:
- \usepackage{eso-pic,graphicx,transparent}
- \usepackage{graphicx}
- \usepackage{fancyhdr}
- \pagestyle{fancy}
- \setlength\headheight{22pt}
- \fancyfoot[RO]{Crossover study design}
- \usepackage{lastpage}
- \cfoot{Page \thepage\ of \pageref{LastPage}}
---

\newpage  
\tableofcontents  
\listoffigures
\listoftables
\newpage

```{r set-options, echo=FALSE, cache=FALSE, warning = FALSE}

 

         set.seed(123)
         startTime<-proc.time()
         library(knitr)
         options(width=60)

         #knitr::opts_chunk$set(dev = 'pdf') # helps with plots
         
         opts_chunk$set(comment = "", warning = FALSE, message = FALSE,
                       echo = TRUE, tidy = TRUE, size="tiny",  cache=FALSE,
                       progress=TRUE, tidy.opts=list(width.cutoff=60),
                         fig.width=7, fig.height=3.5,
                       cache.path = 'program_Cache/',
                       fig.path='figure/')
         
        # opts_chunk$set(tidy.opts=list(width.cutoff=80),tidy=TRUE) 
         
        knitr::knit_hooks$set(inline = function(x) {
          knitr:::format_sci(x, 'md')
        })
         
        
        options(scipen=999)  # remove scientific notation
        

```

# Some functions and parameters

```{r personal functions and params  ,echo=FALSE}

   # useful for rounding 
   p2 <- function(x) {formatC(x, format="f", digits=2)}
   p4 <- function(x) {formatC(x, format="f", digits=4)}

   # number of permutations
   sim = 50000


```

## Hardcode in the data from Hills and Armitage Br. J. clin. Pharmac. (1979), 8, 7-20


```{r function}
 
   patient1 <- c(1,3,4,6,7,9,11,13,16,18,19,21,22,24,25,27,28)
   y1 <- c(8,14,8,9,11,3,6,0,13,10,7,13,8,7,9,10,2)
   treatment1 <- rep("Treatment1",length(patient1))
   order <- rep("First",length(treatment1))
   
   
   treatment2 <- rep("Treatment2",length(treatment1))
   order2 <- rep("Second",length(treatment1))
   y2 <- c(5,10,0,7,6,5,0,0,12,2,5,13,10,7,0,6,2)
   
   
   patient2 <- c(2,5,8,10,12,14,15,17,20,23,26,29)
   y1a<-c(12,6,13,8,8,4,8,2,8,9,7,7) 
   treatment2a <- rep("Treatment2",length(patient2))
   order3 <- rep("First",length(patient2))
   
   
   y1b <-c(11,8,9,8,9,8,14,4,13,7,10,6)
   order4 <- rep("Second",length(patient2))
   treatment3a <- rep("Treatment1",length(patient2))
   
   grp1 <- cbind(as.numeric(patient1) , 
                 (as.character(treatment1)), 
                 (as.character(order)), as.numeric(y1))
   
   grp2 <- cbind(as.numeric(patient1) , 
                 (as.character(treatment2)), 
                 (as.character(order2)), as.numeric(y2))
   
   grp3 <- cbind(as.numeric(patient2) , 
                 (as.character(treatment2a)), 
                 (as.character(order3)), as.numeric(y1a))
   
   grp4 <- cbind(as.numeric(patient2) , 
                 (as.character(treatment3a)), 
                 (as.character(order4)), as.numeric(y1b))
   
   
   all <- as.data.frame(rbind(grp1, grp2, grp3, grp4))
   
   all <- plyr::arrange(all, V1, V2)
   
   names(all) <- c("Patient", "Treatment", "Order", "Response")
   
   all$Patient <- as.numeric(as.character(all$Patient))
   all$Response <- as.numeric(as.character(all$Response))
   
   all <- plyr::arrange(all, Patient, Treatment)
   
  
    knitr::kable(all)
 

```

\newpage

## Plot the data

```{r plot0 , dpi = 400,  fig.height=6,  eval=TRUE} 


    #  order ABBA
    
    d1 <- all
    require(reshape)
    d1$grp <- paste(d1$Treatment, d1$Order, sep=" ")
    d1 <- rename(d1, c(Response="count"))
    d1 <- rename(d1, c(grp="spray"))

    # https://stackoverflow.com/questions/17031039/how-to-sort-a-character-vector-according-to-a-specific-order
    ord <-  c("Treatment1 First", "Treatment2 Second", "Treatment2 First", "Treatment1 Second")
    d1$spray <- factor(d1$spray,levels=ord)
    d1 <- d1[order(d1$spray),]
    
    # create jitter vars
    d1$xj <- cumsum(c(TRUE, d1$spray[-1]!=d1$spray[-length(d1$spray)]))
    d1$xj <- d1$xj+rnorm(nrow(d1), 0, .10)
    d1$yj <- d1$count +  rnorm(nrow(d1), 0, .15) 
    
    
    
    attach(d1)
    
    sprayTypes <- unique(spray)
    
    y <- as.numeric(as.character(d1[,4]))
    
    plot(y, as.factor(d1[,5]) ,  ylim=c(min(y)-0,max(y)+0) , xlim=c(0.5, 4.5) , xaxt="n",
         main="Hall and Armitage data. Plot of AB sequence on left, BA on right\n Dashed lines join paired data, Mean and 95% CI shown",
         xlab="", ylab="Outcome measure", frame.plot  =F , col="white")  
    ##if not white I get a nasty boxplot
    
    axis(1,  at=1:4, labels=F )
    
    text(x=1:length(sprayTypes)*1.15, par("usr")[3]-.8, labels = sprayTypes, 
         srt = 15, pos = 2, xpd = TRUE,  
         cex.axis=.5 )
    
    ##make colours
    chk <- as.character(d1$spray)
    x <- as.data.frame(table(chk))
    
    clr <- c("blue","red")
    
    

    for (i in 1 : length(sprayTypes)) {
      
      n <- sum(spray == sprayTypes[i])
      
      y <- count[spray == sprayTypes[i]]  
      
      xi <- xj[spray == sprayTypes[i]]
      
      yi <- yj[spray == sprayTypes[i]]
  
      points(x=xi, y=yi, pch = 16, cex = .9, 
             col =ifelse(i %in% c(1,4),clr[1],clr[2])
        )
       
      # y is used here
      lines(i +  c(.12, .28), rep(mean(y), 2), lwd = 1, col="black") # 
      
      lines(rep(i + .2, 2),  # start and end
            mean(y) + c(-1.96, 1.96) * sd(y) / sqrt(n),  lwd = 1, col="black"
      ) # vertical
      
    }
    
 
    # join the pairs
    # manage the data , create a new wide data set of 4 coordinates
    
     w1 <- d1[,c(1,3,6,7)]
    
     data1 <- melt(w1 , id.vars = c("Patient", "Order"))
    
     data1$temp <- paste(data1$Order, data1$variable)
     
     wx <- reshape(data1[,c(1,4,5)], idvar="Patient",
                   v.names="value", timevar=c("temp"),
                   direction="wide")
     
     wx <- wx[,c(1,2,4,3,5)]
    
     names(wx) <- c("Patient","x1","y1","x2","y2")
     
    
    for(s in 1:nrow(wx)) {
   
        segments(wx$x1[s], wx$y1[s], wx$x2[s], wx$y2[s], col="brown", lwd=0.5, lty=2)
        
    }
     
```

## Comment


```{r comment1,  echo=FALSE, results='asis'}    


 cat("\nNice plot showing all the data, the blocking in the experiment (patient), the between patient distribution and the sequences of treatment. We would ideally expect AB and BA to be a relection of each other. We show the mean and 95% CI. In fact the standard errors or confidence intervals for individual means are based on an assumption of simple random sampling that is not the case in the crossover trial. So maybe a boxplot would be better. Random noise (jitter) added to datapoints in both x and y axes to make visualisation better.\n") 
   

```

\newpage

## Plot the data once again

```{r plot2}
      
    mydata <- all
    require("PairedData")
    # useful concatenation for plotting
    mydata$temp <- paste(mydata$Treatment, mydata$Order, sep=".")
    
    Treatment1.First <- subset(mydata,  temp == "Treatment1.First", Response, drop = TRUE)
    Treatment2.Second <- subset(mydata,  temp == "Treatment2.Second", Response, drop = TRUE)
     
    pd <- paired(Treatment1.First, Treatment2.Second)
    plot(pd, type = "profile") + theme_bw()
       
    cat("\n\n\\pagebreak\n")
 
    Treatment2.First <- subset(mydata,  temp == "Treatment2.First", Response, drop = TRUE)
    Treatment1.Second <- subset(mydata,  temp == "Treatment1.Second", Response, drop = TRUE)
     
    pd <- paired(Treatment2.First, Treatment1.Second)
   # plot(pd, type = "BA") + theme_bw()
    plot(pd, type = "profile") + theme_bw() 
      
      
```

\newpage

## Period and treatment summary stats, agree with paper

```{r analyse}

 #with(all, tapply(Response, list(Treatment, Order),  mean))

  require(tidyverse)
  all %>% group_by(Treatment, Order) %>%
      summarise_each(funs(n=length(!is.na(.)),
                          mean, 
                          sd, 
                          se=sd(.)/sqrt(n())), 
                     Response)
  
   # calc difference in treatements and summarise
   w <- spread(select (all,-c(Order)), Treatment, Response)
   
   w <-   w %>%  select(Treatment1 , Treatment2) %>% mutate(Response= Treatment1 - Treatment2) #%>% head()
    
   w %>%  summarise(mean = mean(Response),
                      sd = sd(Response),
                       n = length(!is.na(Response)),
                       se = sd/sqrt(n)
       )
    
```

\newpage

## Fit a random effects model (no order nor interaction term)  

```{r nlma}

   require(nlme)
   f <- lme(Response ~ Treatment  , random=~1 | Patient, data=all, na.action="na.omit" )
   anova(f)
   summary(f)$tTable
   intervals(f)
   qqnorm(resid(f), main="Normal Q-Q Plot")
   qqline(resid(f), col="red")
   
   # collect the treatment effect estimate to make inferences later
   z <- as.matrix(summary(f)$tTable)
   Treatment    <- z[2,1][[1]]

```

\newpage

## Fit a linear regression model

```{r lm}      
   
   f <- lm(Response ~ Treatment , data=all, na.action="na.omit" )
   summary(f)
   
   qqnorm(resid(f), main="Normal Q-Q Plot")
   qqline(resid(f), col="red")

```

\newpage

## Function to analyse using permutation approach to duplicate Stephen Senn talk

```{r analyse2}

   Dq1 <- all
   library(data.table)
   Dq1 <- as.data.table(all)
 
   # function to get permuted distribution of treatment effect
   
   perm.dist <- function (block="yes", n.sims=10000){
      
      #set up an array to store parameter estimates
      estArray <- array(NA, dim=c(n.sims,4))
 
      for (s in 1:n.sims){
         
         #~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
         # permute
         
         if (block=="yes") {
            
         # permute within person
         permz <- Dq1[, y := sample(Response), by = Patient] 
         
             } else {
            
         # no blocking
         permz <- Dq1[, y := sample(Response) ]   
         
         }
         
         #~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
         # analysis
         
         # respecting blocking
         possibleError <-  
            tryCatch(f1<-lme( y ~ Treatment, random=~1 | Patient,
                              data=permz,
                              method="REML") , 
                     error=function(e) e)
         
         # http://stackoverflow.com/questions/8093914/skip-to-next-value-of-loop-upon-error-in-r-trycatch
         
         if(!inherits(possibleError, "error")){
            
            modelint <- possibleError
            z <- as.matrix(summary(modelint)$tTable)
            
         }
            
         
         # ignoring blocking
            possibleError2 <-  
               tryCatch(f0<-lm(y ~ Treatment , data=permz) , 
                        error=function(e) e)  
            
            if(!inherits(possibleError, "error")){  
               
               modelint1 <- possibleError2
               zz <- as.matrix(summary(modelint1)$coefficients)
            
            
      }
             
        
         estArray[s,1]  <- z[2,1][[1]]         # collect trt effect estimate
         estArray[s,2]   <- vcov(modelint)[2,2] # collect variance of trt effect estimate
         
         
         estArray[s,3]   <- zz[2,1][[1]]         # collect trt effect estimate
         estArray[s,4]   <- vcov(modelint1)[2,2] # collect variance of trt effect estimate
         
                  }
      
      list(estArray=estArray )
   } 
   
   
```

## Comment


```{r comment2,  echo=FALSE, results='asis'}    



 cat("\nFunction that allows one to permutate data either conditioning on patient (block) correctly by permuting the two values within patient or ignoring the blocking and permuting across all the data. Secondly to run the correct analysis conditioning on patient (LMM model) or not conditioning on patient OLS model.\n") 
  
   
```

\newpage

## Execute the simulations

```{r analyse3}

   block    <- perm.dist(block="yes", n.sims=sim)
   
   no.block <- perm.dist(block="no", n.sims=sim)

```

\newpage

## Plot the treatment effect permutation distributions and present the treatment effect estimate

```{r sim data, eval=TRUE}

   
     A <- block$estArray[,1]
     B <- no.block$estArray[,1] 
    
     plot(density(A), col="red", main="Estimated Treatment Effect (green circle) \nred denotes not blocking, blue conditioned on patient\n", xlab="Permuted treatment effect") 
     lines(density(B),col="blue")
     points(x= Treatment, y=0, col = "darkgreen", cex = 1.5, pch=19)
    
    
     
```

\newpage

## Permutation p values

```{r plot1, eval=TRUE, results='asis'}    
    
     # see Senn 34.09mins right panel youtube, good match!
     sum(abs(block$estArray[,1]) >= abs(Treatment))/sim  # Senn 0.0014
     sum(abs(no.block$estArray[,1]) >= abs(Treatment))/sim  # Senn 0.034
    
```

\newpage

## Summary statisitcs from the permutation approaches

```{r plot2 code,   eval=TRUE} 




    # Manage the estimates
    # Blocked
    A <- as.data.frame( apply(block$estArray, 2, summary))
    names(A) <-c("LMM Mean trt effect","LMM Mean Var of trt effect","Ols Mean trt effect","Ols Mean Var of trt effect")
 
 
     B <- t(as.data.frame(unlist( apply(block$estArray, 2, var))))
     rownames(B) <- NULL
     B <- as.data.frame(B)
     names(B) <-c("Var of LMM trt effect","Var of LMM Var of trt effect","Var of Ols trt effect","Var of Ols Var of trt effect")  
    
     # not blocked
     C <- as.data.frame( apply(no.block$estArray, 2, summary))
     names(C) <- names(A)
 
 
     D <- t(as.data.frame(unlist( apply(no.block$estArray, 2, var))))
     rownames(D) <- NULL
     D <- as.data.frame(D)
     names(D) <-names(B)  
     
     
     
```

\newpage

## Blocked estimates

```{r blocked,   eval=TRUE} 



    knitr::kable(A)
    knitr::kable(B)


```

## Not conditioned on blocking, estimates

```{r not blocked,   eval=TRUE} 


    knitr::kable(C)
    knitr::kable(D)


```

\newpage

## Summary


```{r comment,  echo=FALSE, results='asis'}    

# TABLE 4 COL 3/ table 3 col 4  D[3]/C[4,4]
# table 2 COL 1/ TABLE 1 COL2 OK  B[1] A[4,2]
# TABL 2 COL 3/TABL1 COL 4        B[3] A[4,4]

cat("\nAs Stephen Senn explains in the video. Analysis must relect the block structure. Looking at the distributions of the permutated treatment effects, the effect is the same average difference. The distribution accounting for the fact the same patient is treated on two different occasions is narrower than the distribution that does not condition on patient, in consequence much more unusal event compared to the permutation distribution. (If covariates differ greatly from one patient to another we will see it in the residual error term and we make a judgement of efficacy of something compared to residual error term. Notice the different permutation p-values and compare to parametric p-values. Moral, more important than deciding to use a linear model or permutation test is to condition on block structure of experiment.\n")      

cat("\nWhat happens when you balance but don't condition: That is to say, permute values respecting the fact that they come from a crossover but analysing them as if they came from parallel group trial:\n")

cat("\nApproach, completely randomised and analysed as such", D[3][[1]],". Variance of the treatment effect is equal to the mean of the variance (of treatment effect over all randomisations)",C[4,4][[1]],"\n")

cat("\nApproach, randomised within patient and analysed as such", B[1][[1]],". Variance of the treatment effect is equal to the mean of the variance (of treatment effect over all randomisations)",A[4,2][[1]],"\n")

cat("\nApproach, randomised within patient and analysed as completely randomised", B[3][[1]],". Variance of the treatment effect is not equal to the mean of the variance (of treatment effect over all randomisations)",A[4,4][[1]],"\n")
```

\newpage  

## Computing Environment

```{r}

   sessionInfo()

```

```{r echo=FALSE}

   stopTime<-proc.time()

```

This took `r (stopTime-startTime)[1][[1]]` seconds to execute.

 